Mortar-based entropy stable discontinuous Galerkin methods on non-conforming meshes

### 2019-12-20 docs/ESDG_mortar/main.tex
updating paper

\author{Jesse Chan, David C.\ Del Rey Fernandez, Mario Bencomo}
\title{Mortar-based entropy stable discontinuous Galerkin methods on non-conforming quadrilateral and hexahedral meshes}
\titlerunning{Mortar-based entropy stable DG methods}

@@

\begin{enumerate}
\item Intro
\item Gauss collocation schemes, conditions for entropy stability
\item Increased cost on non-conforming meshes
\item Mortar scheme and entropy stability
\item Analysis of accuracy of Global SBP operator.  Gauss nodes = exact for degree $N$, while Lobatto = exact for degree $(N-1)$.  Approach in Friedrichs + DCDR for $hp$ meshes yields degree $(N-1)$ global operator.
\item Extensions
\begin{enumerate}
\item Curved meshes
\item $p$-nonconforming meshes
\item General elements and why you'd want to implement this.
\end{enumerate}
\item Numerical experiments (2D vortex, 2D entropy conservation test, 3D convergence).  Note that reduced rates of convergence are not observed for $N>1$ on $p$-nonconforming meshes, or on hybrid meshes with mixed quadrature.
\end{enumerate}

\section{Introduction}

%\note{Boilerplate introduction on high order + stability}

Discretely entropy stability has emerged as a methodology for designing high order schemes for nonlinear conservation laws.  Entropy stable discretizations ensure the satisfaction of a semi-discrete entropy inequality by combining specific finite volume numerical fluxes with summation-by-parts (SBP) discretization matrices.  Compared to traditional high order methods, the resulting schemes demonstrate significantly improved robustness in the presence of under-resolved solution features such as shocks or turbulence while retaining high order accuracy.

@@

We can avoid this issue by introducing ``hybridized'' (or decoupled) SBP operators \cite{chan2017discretely, chenreview}, which are defined as the block matrix $\bm{Q}_h$ \cite{chan2019skew}
\[
\bm{Q}_h = \frac{1}{2}\begin{bmatrix}
\bm{Q}-\bm{Q}^T & \bm{E}^T\bm{B}\\
-\bm{B}\bm{E} & \bm{B}
\end{bmatrix}.
\]
Note that the hybridized SBP operator satisfies a block form of the SBP property
\begin{equation}
\bm{Q}_h + \bm{Q}_h^T = \begin{bmatrix}
\bm{0} & \\
& \bm{B}\end{bmatrix}.
\label{eq:hsbp}
\end{equation}
where $\bm{E}$ does not appear in the block boundary matrix on the right hand side.  Additionally, it is straightforward to show that 
\[
\bm{Q}_h\bm{1} = \bm{0}
\]
using the GSBP property and properties of $\bm{Q}$.

The aforementioned matrices can now be used to construct high order accurate entropy stable and entropy conservative schemes.  We first introduce an entropy conservative numerical flux \cite{tadmor1987numerical}.  Let $\bm{u}_L, \bm{u}_R$ denote left and right states in the conservative variables.  Then, an entropy conservative numerical flux is a vector-valued function $\bm{f}_S$ which satisfies the following three properties
\begin{gather*}
\bm{f}_S\LRp{\bm{u},\bm{u}} = \bm{f}(\bm{u}), \qquad \text{(consistency)}\\
\bm{f}_S\LRp{\bm{u}_L,\bm{u}_R} = \bm{f}_S(\bm{u}_R,\bm{u}_L), \qquad \text{(symmetry)}\\
\LRp{\bm{v}_L-\bm{v}_R}^T\bm{f}_S\LRp{\bm{u}_L,\bm{u}_R} = \psi(\bm{u}_L) - \psi(\bm{u}_R), \qquad \text{(conservation)}.
\end{gather*}
Here, $\bm{v}_L, \bm{v}_R$ are the entropy variables evaluated at $\bm{u}_L, \bm{u}_R$, and $\psi$ is the entropy potential which appears in (\ref{eq:weakentropyineq}).  Then, an entropy conservative scheme on the reference interval is given as
\begin{gather}
\bm{M}\td{\bm{u}_h}{t} + \begin{bmatrix} \bm{I} \\ \bm{E} \end{bmatrix}^T
\LRp{\bm{Q}_h \circ \bm{F}}\bm{1} + \bm{E}^T\bm{B}\LRp{\bm{f}^*-\bm{f}(\tilde{\bm{u}}_f)} = 0 \label{eq:esdg1D}\\ 
\tilde{\bm{u}} = \begin{bmatrix}
\bm{u}_h\\
\tilde{\bm{u}}_f
\end{bmatrix}, \qquad \tilde{\bm{u}}_f = \bm{u}\LRp{\bm{E}\bm{v}(\bm{u}_h)}, \qquad \bm{f}^* = \bm{f}_S\LRp{\tilde{\bm{u}}_f^+,\tilde{\bm{u}}_f}.
\nonumber
\end{gather}
Here, $\tilde{\bm{u}}_f^+$ denotes some exterior state which can be used to enforce boundary conditions or coupling between different elements, and $\circ$ denotes the matrix Hadamard product.

\note{Maybe add proof here and show how rest of proofs just rely on SBP and $\bm{Q}_h\bm{1} = \bm{0}$?}.  For proofs of high order accuracy and entropy conservation, we refer the reader to \cite{chan2018efficient}.

The matrix $\bm{F}$ is formed by computing ${\bm{F}}_{ij} = \bm{f}_S\LRp{\tilde{\bm{u}}_i,\tilde{\bm{u}}_j}$ on the fly.  Each entry requires evaluating the entropy conservative numerical flux $\bm{f}_S$, resulting in a large number of floating point operations \cite{wintermeyer2018entropy}.  However, due to properties of the Hadamard product in (\ref{eq:esdg1D}), flux evaluations are only necessary between pairs of nodes corresponding to non-zero entries of $\bm{Q}_h$.  We will revisit this point later in the construction of entropy stable schemes on non-conforming meshes.

@@

The formulation (\ref{eq:weakentropyineq}) can be extended to higher dimensions through a tensor product construction.  For simplicity, we illustrate this in 2D (the extension to 3D is straightforward).  Let $\bm{M}_{\rm 1D}, \bm{Q}_{\rm 1D}$ denote one-dimensional SBP norm (mass) and differentiation matrices, and let $\bm{E}_{\rm 1D}$ denote the 1D face interpolation matrix.  We define multi-dimensional  mass and differentiation matrices in terms of Kronecker products  
\[
\hat{\bm{Q}}_1 = \bm{Q}_{\rm 1D} \otimes \bm{M}_{\rm 1D}, \qquad \hat{\bm{Q}}_2  = \bm{M}_{\rm 1D} \otimes \bm{Q}_{\rm 1D}, \qquad \hat{\bm{M}} = \bm{M}_{\rm 1D} \otimes  \bm{M}_{\rm 1D}.
\]
We also construct 2D face interpolation matrices from Kronecker products.  Let $\bm{E}_{\rm 1D}$ denote the one-dimensional face interpolation matrix, and let $\bm{B}_{\rm 1D}$ denote the 1D boundary matrix in (\ref{eq:gsbp}).  For a specific ordering of the face points, the two-dimensional face interpolation matrix $\bm{E}$ and reference boundary matrices $\hat{\bm{B}}_1, \hat{\bm{B}}_2$ are given by
\begin{equation}
\bm{E} = \begin{bmatrix}
\bm{E}_{\rm 1D} \otimes \bm{I}_{2}\\
\bm{I}_{2} \otimes \bm{E}_{\rm 1D} 
\end{bmatrix}, \qquad 
\hat{\bm{B}}_1 = \begin{bmatrix}
\bm{B}_{\rm 1D} \otimes \bm{M}_{\rm 1D} & \\
& \bm{0}
\end{bmatrix}, \qquad 
\hat{\bm{B}}_2 = \begin{bmatrix}
\bm{0} &\\
& \bm{M}_{\rm 1D} \otimes \bm{B}_{\rm 1D} 
\end{bmatrix},
\label{eq:EBdefs}
\end{equation}
where $\bm{I}_2$ is the $2\times 2$ identity matrix.  Recall that $\bm{M}_{\rm 1D}$ is the diagonal matrix of quadrature weights, so $\hat{\bm{B}}_i$ are diagonal.

The 2D differentiation and interpolation matrices $\hat{\bm{Q}}_i, \bm{E}$ can now be used to construct 2D hybridized SBP operators.  Let $\hat{\bm{Q}}_{i,h}$ denote the hybridized SBP operator for the $i$th coordinate on the reference element, where $\hat{\bm{Q}}_{i,h}$ is defined as
\begin{align*}
\hat{\bm{Q}}_{i,h} = \frac{1}{2}\begin{bmatrix}
\hat{\bm{Q}}_i - \hat{\bm{Q}}_i^T & \bm{E}^T\hat{\bm{B}}_i\\
-\hat{\bm{B}}_i\bm{E} & \hat{\bm{B}}_i
\end{bmatrix}.
\end{align*}
These matrices satisfy the multi-dimensional SBP property
\begin{align}
\hat{\bm{Q}}_{i,h} + \hat{\bm{Q}}_{i,h}^T = \begin{bmatrix}
\bm{0} & \\
& \hat{\bm{B}}_i
\end{bmatrix}.
\label{eq:hsbp}
\end{align}
The construction of 3D differentiation and interpolation matrices proceeds similarly.  Then, a $d$-dimensional entropy conservative formulation on the reference element $[-1,1]^d$ is given as 
\begin{gather}
\hat{\bm{M}}\td{\bm{u}_h}{t} + \begin{bmatrix} \bm{I} \\ \bm{E} \end{bmatrix}^T
\sum_{i=1}^d \LRp{\hat{\bm{Q}}_{i,h} \circ \bm{F}_i}\bm{1} + \bm{E}^T\hat{\bm{B}}_i\LRp{\bm{f}_i^*-\bm{f}_i(\tilde{\bm{u}}_f)} = 0 \label{eq:esdgref}\\
\LRp{\bm{F}_i}_{jk} = \bm{f}^i_S\LRp{\tilde{\bm{u}}_j,\tilde{\bm{u}}_k}
\end{gather}
where $\bm{f}^i_S$ is the numerical flux for the $i$th coordinate direction, and $\tilde{\bm{u}},\tilde{\bm{u}}_f$ are defined as in (\ref{eq:esdg1D}).

Finally, we can extend this formulation to the multi-element case.  Suppose the domain $\Omega$ is decomposed into non-overlapping elements $D^k$ which are images of the reference mapping, such that $D^k$ is a differentiable mapping of the reference element $\hat{D} = [-1,1]^d$.  Let $\bm{x}_i$ denote the $i$th physical coordinate on $D^k$ and let $\hat{\bm{x}}_j$ denote the $j$th reference coordinate.  The mapping between reference and physical element induces scaled geometric terms 
\[
G_{ij} = J\pd{\hat{\bm{x}}_j}{\bm{x}_i},
\]
where $J$ denotes the determinant of the Jacobian of the physical-to-reference mapping.  We refer to $J$ as the Jacobian from here onwards.
These geometric terms also relate the normal vectors on reference and physical elements.  Let $\hat{\bm{n}}\hat{J}_f$ denote the outward normal vector on $\hat{D}$, scaled by the reference face Jacobians.  Then, the (scaled) physical normal vectors on $D^k$ are related to $\hat{\bm{n}}\hat{J}_f$ through
\[
\bm{n}J_f = G\hat{\bm{n}}\hat{J}_f.
\]
We assume that the mesh is watertight, such that the scaled normal vectors are equal and opposite across shared faces between two elements.

We can now assemble physical mass and differentiation matrices on $D^k$ from reference mass and differentiation matrices using the chain rule.  Let $\hat{\bm{Q}}_{j,h}$ denote the $j$th reference differentiation matrix on $\hat{D}$.  We can define the physical differentiation matrices $\bm{Q}_{i,h}$ on $D^k$ via the skew-symmetric splitting
\[
\bm{Q}_{i,h} = \frac{1}{2}\sum_{j=1}^d \diag{\bm{g}_{ij}}\hat{\bm{Q}}_{j,h} + \hat{\bm{Q}}_{j,h} \diag{\bm{g}_{ij}},
\]
where $\bm{g}_{ij}$ denotes the vector of values of $G_{ij} = J\pd{\hat{\bm{x}}_j}{\bm{x}_i}$ at both volume and surface points.  We also assume that the geometric terms $\bm{g}_{ij}$ are constructed such that they satisfy a discrete geometric conservation law (GCL)
\[
\sum_{j=1}^d \hat{\bm{Q}}_{j,h} \bm{g}_{ij} = \bm{0}.
\]
The discrete GCL is satisfied automatically for isoparametric mappings in 2D, and there exist several techniques to enforce the satisfaction of a discrete GCL on various three-dimensional domains \cite{thomas1979geometric, kopriva2006metric, crean2018entropy, chan2018discretely, kozdon2018energy, kopriva2019free}.

Because the reference matrices $\hat{\bm{Q}}_{j,h}$ satisfy the SBP property (\ref{eq:hsbp}), one can show \cite{chan2018discretely} that the physical matrices $\bm{Q}_{i,h}$ satisfy a physical SBP property.  
\begin{lemma}
\label{lemma:Qhprops}

\[
\bm{Q}_{i,h} + \bm{Q}_{i,h}^T = \begin{bmatrix}
\bm{0} &\\
& \bm{B}_i \end{bmatrix}, \qquad \bm{B}_i = \diag{\bm{n}_i \circ \bm{w}_f},
\]
where $\bm{n}_i$ is a vector whose entries consist of values of the $i$th component of the scaled normals $n_iJ_f$ at face points, and $\bm{w}_f$ is the vector of face quadrature weights $w_i$.  
\end{lemma}

Then, an entropy conservative scheme is given by
\begin{gather}
{\bm{M}}\td{\bm{u}_h}{t} + \begin{bmatrix} \bm{I} \\ \bm{E} \end{bmatrix}^T
\sum_{i=1}^d \LRp{{\bm{Q}}_{i,h} \circ \bm{F}_i}\bm{1} + \bm{E}^T{\bm{B}}_i\LRp{\bm{f}_i^*-\bm{f}_i(\tilde{\bm{u}}_f)} = 0 \label{eq:esdg}
\end{gather}
where $\bm{F}_i$ is defined as in (\ref{eq:esdgref}).  Here, the exterior state $\tilde{\bm{u}}_f$ used to evaluate the numerical flux $\bm{f}_i^*$ corresponds either to interface values on a neighboring element or an exterior state used to enforce boundary conditions.

>>>>

\author{Jesse Chan, Mario Bencomo, David C.\ Del Rey Fernandez}
\title{Mortar-based entropy stable discontinuous Galerkin methods on non-conforming meshes}
\titlerunning{Mortar-based entropy stable DG methods}

@@

\section{Introduction}
Discretely entropy stability has emerged as a methodology for designing high order schemes for nonlinear conservation laws.  Entropy stable discretizations ensure the satisfaction of a semi-discrete entropy inequality by combining specific finite volume numerical fluxes with summation-by-parts (SBP) discretization matrices.  Compared to traditional high order methods, the resulting schemes demonstrate significantly improved robustness in the presence of under-resolved solution features such as shocks or turbulence while retaining high order accuracy.

@@

We can avoid this issue by introducing ``hybridized'' (or decoupled) SBP operators \cite{chan2017discretely, chenreview}, which are defined as the block matrix $\bm{Q}_h$ \cite{chan2019skew}
\[
\bm{Q}_h = \frac{1}{2}\begin{bmatrix}
\bm{Q}-\bm{Q}^T & \bm{E}^T\bm{B}\\
-\bm{B}\bm{E} & \bm{B}
\end{bmatrix}.
\]
Note that the hybridized SBP operator satisfies a block form of the SBP property
\begin{equation}
\bm{Q}_h + \bm{Q}_h^T = \bm{B}_h, \qquad \bm{B}_h =  \begin{bmatrix}
\bm{0} & \\
& \bm{B}\end{bmatrix}.
\label{eq:hsbp}
\end{equation}
where $\bm{E}$ does not appear in the block boundary matrix on the right hand side.  Here, we have used $\bm{0}$ to denote a matrix or vector of zeros, where the size is inferred from context.  We will adopt a similar convention for the vector of ones $\bm{1}$.

The matrix $\bm{Q}_h$ also satisfies
\[
\bm{Q}_h\bm{1} = 
\frac{1}{2}\begin{bmatrix}
\bm{Q}\bm{1}-\bm{Q}^T\bm{1} + \bm{E}^T\bm{B}\bm{1}\\
-\bm{B}\bm{E}\bm{1} - \bm{B}\bm{1}
\end{bmatrix}= 
\frac{1}{2}\begin{bmatrix}
-\bm{Q}^T\bm{1} + \bm{E}^T\bm{B}\bm{1}\\
\bm{0}
\end{bmatrix}
\]
where we have used that $\bm{E}\bm{1} = \bm{1}$ (since $\bm{E}$ is a high order accurate boundary interpolation matrix) and $\bm{Q}\bm{1} = \bm{0}$ (since $\bm{Q}$ is a differentiation matrix).  Combining the GSBP property $-\bm{Q}^T + \bm{E}^T\bm{B}\bm{E} = \bm{Q}$ and $\bm{E}\bm{1} = \bm{1}$ then yields that 
\begin{equation}
\bm{Q}_h\bm{1} = 
\frac{1}{2}\begin{bmatrix}
-\bm{Q}^T\bm{1} + \bm{E}^T\bm{B}\bm{1}\\
\bm{0}
\end{bmatrix} = 
\frac{1}{2}\begin{bmatrix}
\bm{Q}\bm{1}\\
\bm{0}
\end{bmatrix} 
= \frac{1}{2}\begin{bmatrix}
\bm{0}\\
\bm{0}
\end{bmatrix} = \bm{0}.
\label{eq:Qh1}
\end{equation}

The aforementioned matrices can now be used to construct high order accurate entropy stable and entropy conservative schemes.  We first introduce an entropy conservative numerical flux \cite{tadmor1987numerical}.  Let $\bm{u}_L, \bm{u}_R$ denote left and right states in the conservative variables.  Then, an entropy conservative numerical flux is a vector-valued function $\bm{f}_S$ which satisfies the following three properties
\begin{gather*}
\bm{f}_S\LRp{\bm{u},\bm{u}} = \bm{f}(\bm{u}), \qquad \text{(consistency)}\\
\bm{f}_S\LRp{\bm{u}_L,\bm{u}_R} = \bm{f}_S(\bm{u}_R,\bm{u}_L), \qquad \text{(symmetry)}\\
\LRp{\bm{v}_L-\bm{v}_R}^T\bm{f}_S\LRp{\bm{u}_L,\bm{u}_R} = \psi(\bm{u}_L) - \psi(\bm{u}_R), \qquad \text{(conservation)}.
\end{gather*}
Here, $\bm{v}_L, \bm{v}_R$ are the entropy variables evaluated at $\bm{u}_L, \bm{u}_R$, and $\psi$ is the entropy potential which appears in (\ref{eq:weakentropyineq}).  Then, an entropy conservative scheme on the reference interval is given as
\begin{gather}
\bm{M}\td{\bm{u}_h}{t} + \begin{bmatrix} \bm{I} \\ \bm{E} \end{bmatrix}^T
\LRp{2\bm{Q}_h \circ \bm{F}}\bm{1} + \bm{E}^T\bm{B}\LRp{\bm{f}^*-\bm{f}(\tilde{\bm{u}}_f)} = 0 \label{eq:esdg1D}\\ 
\tilde{\bm{u}} = \begin{bmatrix}
\bm{u}_h\\
\tilde{\bm{u}}_f
\end{bmatrix}, \qquad \tilde{\bm{u}}_f = \bm{u}\LRp{\bm{E}\bm{v}(\bm{u}_h)}, \qquad \bm{f}^* = \bm{f}_S\LRp{\tilde{\bm{u}}_f^+,\tilde{\bm{u}}_f}.
\nonumber
\end{gather}
Here, $\tilde{\bm{u}}_f^+$ denotes some exterior state which can be used to enforce boundary conditions or coupling between different elements, and $\circ$ denotes the matrix Hadamard product.

\begin{theorem}
Assuming continuity in time, the formulation (\ref{eq:esdg1D}) is entropy conservative in the sense that
\[
\bm{1}^T\bm{M}\td{S(\bm{u}_h)}{t} + \bm{1}^T\bm{B}\LRp{\bm{v}_f^T\bm{f}^* - \psi(\tilde{\bm{u}}_f)} = 0, \qquad \bm{v}_f =\bm{E}\bm{v}(\bm{u}_h).
\]
\end{theorem}
\begin{proof}
The proof uses similar techniques as proofs in other papers \cite{chen2017entropy, crean2018entropy, chan2017discretely, chan2019skew}, and we repeat it for completeness.  We begin by testing (\ref{eq:esdg1D}) with the vector of entropy variables evaluated at nodal points $\bm{v}(\bm{u}_h)$.  We apply the chain rule in time and use that $\bm{M}$ is diagonal to show that 
\[
\bm{v}(\bm{u}_h)^T\bm{M}\td{\bm{u}}{t} = \bm{1}^T\bm{M}\LRp{\bm{v}(\bm{u}_h)\circ \td{\bm{u}_h}{t}} =  \bm{1}^T\bm{M}\td{S(\bm{u}_h)}{t}.
\]
For the volume term, we use the SBP property to arrive at
\begin{align*}
\bm{v}(\bm{u}_h)^T\begin{bmatrix} \bm{I} \\ \bm{E} \end{bmatrix}^T\LRp{2\bm{Q}_h \circ \bm{F}}\bm{1} &= \tilde{\bm{v}}^T\LRp{\LRp{\bm{Q}_h-\bm{Q}_h^T} \circ \bm{F}}\bm{1}  + \LRp{\bm{E}\bm{v}(\bm{u}_h)}^T\LRp{\bm{B}_h \circ\bm{F}}\bm{1},
\end{align*}
where we have introduced $\tilde{\bm{v}} = \LRp{\begin{bmatrix} \bm{I} \\ \bm{E} \end{bmatrix}\bm{v}(\bm{u}_h)}$.  
Note that, by the consistency of the entropy conservative flux $\bm{f}_S$, the diagonal of $\bm{F}$ is the vector $\bm{f}(\tilde{\bm{u}}_f)$.  Using that $\bm{B}$ is diagonal, the latter term reduces to 
\[
\LRp{\bm{E}\bm{v}(\bm{u}_h)}^T\LRp{\bm{B}_h \circ\bm{F}}\bm{1} = \bm{v}_f^T \bm{B}\bm{f}(\tilde{\bm{u}}_f) =  \bm{1}^T \bm{B}\LRp{\bm{v}_f^T\bm{f}(\tilde{\bm{u}}_f)}.
\]
The final part of the proof uses that 
\begin{align*}
\tilde{\bm{v}}^T\LRp{\LRp{\bm{Q}_h-\bm{Q}_h^T} \circ \bm{F}}\bm{1} &= \sum_j {\tilde{\bm{v}}}_j \LRp{\bm{Q}_h-\bm{Q}_h^T}_{ij} \bm{F}_{ij}\\
&= \sum_j \LRp{\bm{Q}_h}_{ij} \LRp{{\tilde{\bm{v}}}_i-{\tilde{\bm{v}}}_j}^T\bm{f}_{S}\LRp{\tilde{\bm{u}}_i,\tilde{\bm{u}}_j}\\
&= \sum_j \LRp{\bm{Q}_h}_{ij} \LRp{\psi(\tilde{\bm{u}}_i)-\psi(\tilde{\bm{u}}_j)} = \tilde{\bm{u}}^T\bm{Q}_h\bm{1}- \bm{1}^T\bm{Q}_h\tilde{\bm{u}}
\end{align*}
The term $\bm{Q}_h\bm{1} = \bm{0}$ by (\ref{eq:Qh1}).  The proof is completed by applying the SBP property (\ref{eq:hsbp}) to the remaining term, which reduces to $-\bm{1}^T\bm{B}_h\tilde{\bm{u}}=-\bm{1}^T\bm{B}\tilde{\bm{u}_f}$.
\qed\end{proof}

Note that the proof of entropy conservation for (\ref{eq:esdg1D}) relies mainly on the fact that $\bm{Q}_h$ is conservative (e.g., exact for constants) (\ref{eq:Qh1}) and that $\bm{Q}_h$ satisfies the SBP property (\ref{eq:hsbp}).  Extensions to the multi-element, multi-dimensional, and non-conforming setting will utilize the same properties.

%\note{Maybe add proof here and show how rest of proofs just rely on SBP and $\bm{Q}_h\bm{1} = \bm{0}$?}.  For proofs of high order accuracy and entropy conservation, we refer the reader to \cite{chan2018efficient}.

%The matrix $\bm{F}$ is formed by computing ${\bm{F}}_{ij} = \bm{f}_S\LRp{\tilde{\bm{u}}_i,\tilde{\bm{u}}_j}$ on the fly.  Each entry requires evaluating the entropy conservative numerical flux $\bm{f}_S$, resulting in a large number of floating point operations \cite{wintermeyer2018entropy}.  However, due to properties of the Hadamard product in (\ref{eq:esdg1D}), flux evaluations are only necessary between pairs of nodes corresponding to non-zero entries of $\bm{Q}_h$.  We will revisit this point later in the construction of entropy stable schemes on non-conforming meshes.

@@

The formulation (\ref{eq:weakentropyineq}) can be extended to higher dimensions through a tensor product construction.  For simplicity, we illustrate this in 2D (the extension to 3D is straightforward).  Let $\bm{M}_{\rm 1D}, \bm{Q}_{\rm 1D}$ denote one-dimensional generalized SBP norm (mass) and differentiation matrices, and let $\bm{E}_{\rm 1D}$ denote the 1D face interpolation matrix.  We define multi-dimensional  mass and differentiation matrices in terms of Kronecker products  
\begin{equation}
\hat{\bm{Q}}_1 = \bm{Q}_{\rm 1D} \otimes \bm{M}_{\rm 1D}, \qquad \hat{\bm{Q}}_2  = \bm{M}_{\rm 1D} \otimes \bm{Q}_{\rm 1D}, \qquad \hat{\bm{M}} = \bm{M}_{\rm 1D} \otimes  \bm{M}_{\rm 1D}.
\label{eq:kron}
\end{equation}
We also construct 2D face interpolation matrices from Kronecker products.  Let $\bm{E}_{\rm 1D}$ denote the one-dimensional face interpolation matrix, and let $\bm{B}_{\rm 1D}$ denote the 1D boundary matrix in (\ref{eq:gsbp}).  For a specific ordering of the face points, the two-dimensional face interpolation matrix $\bm{E}$ and reference boundary matrices $\hat{\bm{B}}_1, \hat{\bm{B}}_2$ are given by
\begin{equation}
\bm{E} = \begin{bmatrix}
\bm{E}_{\rm 1D} \otimes \bm{I}_{2}\\
\bm{I}_{2} \otimes \bm{E}_{\rm 1D} 
\end{bmatrix}, \qquad 
\hat{\bm{B}}_1 = \begin{bmatrix}
\bm{B}_{\rm 1D} \otimes \bm{M}_{\rm 1D} & \\
& \bm{0}
\end{bmatrix}, \qquad 
\hat{\bm{B}}_2 = \begin{bmatrix}
\bm{0} &\\
& \bm{M}_{\rm 1D} \otimes \bm{B}_{\rm 1D} 
\end{bmatrix},
\label{eq:EBdefs}
\end{equation}
where $\bm{I}_2$ is the $2\times 2$ identity matrix.  Recall that $\bm{M}_{\rm 1D}$ is the diagonal matrix of quadrature weights, so $\hat{\bm{B}}_i$ are diagonal.

The 2D differentiation and interpolation matrices $\hat{\bm{Q}}_i, \bm{E}$ can now be used to construct 2D hybridized SBP operators.  Let $\hat{\bm{Q}}_{i,h}$ denote the hybridized SBP operator for the $i$th coordinate on the reference element, where $\hat{\bm{Q}}_{i,h}$ is defined as
\begin{align}
\hat{\bm{Q}}_{i,h} = \frac{1}{2}\begin{bmatrix}
\hat{\bm{Q}}_i - \hat{\bm{Q}}_i^T & \bm{E}^T\hat{\bm{B}}_i\\
-\hat{\bm{B}}_i\bm{E} & \hat{\bm{B}}_i
\end{bmatrix}.
\label{eq:hsbpdef}
\end{align}
These matrices satisfy the following multi-dimensional properties:
\begin{lemma}
\label{lemma:hsbprefprops}
Let $\hat{\bm{Q}}_{i}$ be defined as in (\ref{eq:kron}) and let $\hat{\bm{Q}}_{i,h}$ be defined as in (\ref{eq:hsbpdef}).  Then, 
\begin{align}
\hat{\bm{Q}}_{i,h} + \hat{\bm{Q}}_{i,h}^T = \begin{bmatrix}
\bm{0} & \\
& \hat{\bm{B}}_i
\end{bmatrix}, \qquad \hat{\bm{Q}}_{i,h}\bm{1} = \bm{0}.
\end{align}
\end{lemma}
The SBP property follows by construction of $\hat{\bm{Q}}_{i,h}$.  The property $\hat{\bm{Q}}_{i,h}\bm{1} = \bm{0}$ is a consequence of the Kronecker product and the same arguments used to derive (\ref{eq:Qh1}).  The construction of 3D differentiation and interpolation matrices proceeds similarly, and the matrices satisfy the same properties as described in Lemma~\ref{lemma:hsbprefprops}.

Then, a $d$-dimensional entropy conservative formulation on the reference element $[-1,1]^d$ is given as 
\begin{gather}
\hat{\bm{M}}\td{\bm{u}_h}{t} + \begin{bmatrix} \bm{I} \\ \bm{E} \end{bmatrix}^T
\sum_{i=1}^d \LRp{2\hat{\bm{Q}}_{i,h} \circ \bm{F}_i}\bm{1} + \bm{E}^T\hat{\bm{B}}_i\LRp{\bm{f}_i^*-\bm{f}_i(\tilde{\bm{u}}_f)} = 0 \label{eq:esdgref}\\
\LRp{\bm{F}_i}_{jk} = \bm{f}^i_S\LRp{\tilde{\bm{u}}_j,\tilde{\bm{u}}_k}
\end{gather}
where $\bm{f}^i_S$ is the numerical flux for the $i$th coordinate direction, and $\tilde{\bm{u}},\tilde{\bm{u}}_f$ are defined as in (\ref{eq:esdg1D}).

Finally, we can extend this formulation to the multi-element case.  Suppose the domain $\Omega$ is decomposed into non-overlapping elements $D^k$ which are images of the reference mapping, such that $D^k$ is a differentiable mapping of the reference element $\hat{D} = [-1,1]^d$.  Let $\bm{x}_i$ denote the $i$th physical coordinate on $D^k$ and let $\hat{\bm{x}}_j$ denote the $j$th reference coordinate.  The mapping between reference and physical element induces scaled geometric terms 
\[
G_{ij} = J\pd{\hat{\bm{x}}_j}{\bm{x}_i},
\]
where $J$ denotes the determinant of the Jacobian of the physical-to-reference mapping.  We refer to $J$ as the Jacobian from here onwards.
These geometric terms also relate the normal vectors on reference and physical elements.  Let $\hat{\bm{n}}\hat{J}_f$ denote the outward normal vector on $\hat{D}$, scaled by the reference face Jacobians.  Then, the (scaled) physical normal vectors on $D^k$ are related to $\hat{\bm{n}}\hat{J}_f$ through
\begin{equation}
\bm{n}J_f = G\hat{\bm{n}}\hat{J}_f.
\label{eq:nJ_Gnhat}
\end{equation}
We assume that the mesh is watertight, such that the scaled normal vectors are equal and opposite across shared faces between two elements.  We also assume that the geometric terms ${G}_{ij}$ are constructed such that they are polynomials of degree less than or equal to $N$ and satisfy a discrete geometric conservation law (GCL)
\begin{equation}
\sum_{j=1}^d \pd{}{\hat{x}_j} G_{ij} = 0.
%\sum_{j=1}^d \hat{\bm{Q}}_{j,h} \bm{G}_{ij} = \bm{0}.
\label{eq:dgcl}
\end{equation}
The discrete GCL is satisfied automatically for isoparametric mappings in 2D, and there exist several techniques to enforce the satisfaction of a discrete GCL on various three-dimensional domains \cite{thomas1979geometric, kopriva2006metric, crean2018entropy, chan2018discretely, kozdon2018energy, kopriva2019free}.

We can now assemble physical mass and differentiation matrices on $D^k$ from reference mass and differentiation matrices using the chain rule.  Let $\hat{\bm{Q}}_{j,h}$ denote the $j$th reference differentiation matrix on $\hat{D}$.  We can define the physical differentiation matrices $\bm{Q}_{i,h}$ on $D^k$ via the skew-symmetric splitting
\begin{equation}
\bm{Q}_{i,h} = \frac{1}{2}\sum_{j=1}^d \diag{\bm{G}_{ij}}\hat{\bm{Q}}_{j,h} + \hat{\bm{Q}}_{j,h} \diag{\bm{G}_{ij}},
\label{eq:curvedQ}
\end{equation}
where $\bm{G}_{ij}$ denotes the vector of values of $G_{ij} = J\pd{\hat{\bm{x}}_j}{\bm{x}_i}$ at both volume and surface points.  Since $G_{ij}$ is assumed to be a polynomial of degree less than or equal to $N$, $\bm{G}_{ij}$ is constructed using polynomial interpolation.
